---
title: "RMS Titanic"
author: "Navya Tiwari"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

> - `Class`, `Sex`, `Age`, `Survived`, and `n`.

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% 
  summarize(total = sum(n))
```

**Observations**:

- The dataset has 2,201 people, while Wikipedia lists 2,208, so there is a difference of 7 people.
- Are there any differences?
  - Yes there is a difference of 7 people.
  - This question also made me think of how this difference might actually affect the data.
  - Initially I felt that this is a pretty small difference compared to the total number of people aboard.
  -  But, the class that the people actually belong to has the power to greatly influence data, for example we know first class had comparatively lesser people than second and third class if these people were all part of the first class then that might influence or skew any class comparisons we might want to perform with this data
- If yes, what might account for those differences?
  - The difference could come from uncertainty in the historical records.
  - Wikipedia mentions confusion in passenger lists, including people who cancelled at the last minute and passengers traveling under aliases so stowaways as well.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(
    aes(
      x = Class,
      y = n,
      fill = Sex
    )
  )
```

**Observations**:

- On first look something that seems to be a pattern across all classes and crew is that the female survivors were greater in number than male survivors

- Another thing that catches the eye is that the proportion male to female survivors based on tihs graph seems to decrease as we move from first class toward second, third class and then crew.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(
    aes(
      x = Class,
      y = Prop,
      fill = Sex
    )
  )
```

**Observations**:

- Write your observations here.

  - The proportions give a different picture than the raw counts. For example, the large number of surviving male crew members in Q3 does not mean that male crew members had a particularly high survival rate.

  - Female survival proportions generally seem higher than male survival proportions.

- Is there anything *fishy* going on in your plot?

  - Some of the bars go above 1,
  - This happens because there are separate proportions for adults and children, but the graph is stacking them together. These proportions come from different groups, so adding them together does not give one meaningful survival proportion

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(
    aes(
      x = Class,
      y = Prop,
      fill = Sex
    ),
    position = "dodge"
  ) +
  facet_grid(Age ~ .)
```

**Observations**:

- (Write your observations here.)

  - Female adults had much higher survival proportions than male adults in every class.

  - First- and second-class children in this dataset had very high survival proportions, while third-class children had much lower survival proportions.

  - Class also seems related to survival. First and second class women had higher survival proportions than third-class women, and first-class adult men had a higher survival proportion than second- or third-class adult men.

- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.

  - The Q4 graph looked strange because it combined the adult and child proportions. Separating the data by age makes it clear that survival patterns were different for adults and children.

<!-- -->

- "Women and children first policy"

  - Adult women have dramatically higher survival proportions than adult men.

  - Class seems to matter even within the same sex.

  - Third-class children look very different from first and second class children.

  - The crew is especially interesting because its pattern is very different for men versus women.

  - class clearly matters too.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
