---
title: "Michelson Speed-of-light Measurements"
author: "(Navya Tiwari)"
date: 2020-
output: 
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: When studying physical problems, there is an important
distinction between *error* and *uncertainty*. The primary purpose of
this challenge is to dip our toes into these factors by analyzing a real
dataset.

*Reading*: [Experimental Determination of the Velocity of
Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
(Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to
measure the speed of light. His approach was a development upon the
method of Foucault[3], and resulted in a new estimate of
$v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very
close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this
challenge, you will analyze Michelson's original data, and explore some
of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code
chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the
value `LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition)
in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse()
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1
  = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r q1-task}
## TODO: Compute summaries
df_q1 <-
  df_michelson %>%
  group_by(Distinctness) %>%
  summarize(
    n = n(),
    MeanVelocity = mean(Velocity)
  )

df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:

- Write your observations here!

  - At first, I mostly noticed that the distinctness 1 group had a lower
    mean velocity, so I thought poorer image quality might just lead to
    lower measurements. But when I looked more closely, I realized that
    distinctness 2 and 3 are almost identical in their mean velocities,
    even though one is supposed to be “fair” and the other “good.” That
    makes me think the relationship is probably not as simple as “better
    distinctness means a higher or more accurate measurement.” It seems
    more like the distinctness 1 measurements are the ones that really
    stand out, while the other two groups behave pretty similarly.

  - I also noticed that there are only 15 measurements in the
    distinctness 1 group, compared with 39 and 46 in the other groups.
    So I would be a little careful about making a strong conclusion from
    that group alone.

- Why might your table differ from Michelson's?

  - My table is probably a little different because I kept the exact
    calculated means with decimals, while Michelson seems to have
    rounded his reported values. For example, my mean of 299861.7 km/s
    would show up as about 299860 km/s in his table.

The `Velocity` values in the dataset are the speed of light *in air*;
Michelson introduced a couple of adjustments to estimate the speed of
light in a vacuum. In total, he added $+92$ km/s to his mean estimate
for `VelocityVacuum` (from Michelson (1880), pg. 141). While the
following isn't fully rigorous ($+92$ km/s is based on the mean
temperature), we'll simply apply this correction to all the observations
in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <-
  df_michelson %>%
  mutate(
    VelocityVacuum = Velocity + 92
  )

df_q2
```

- I noticed here that after adding the 92 km/s correction, every
  measurement shifts upward by the same amount. This does not change the
  pattern between the measurements, it just changes them from estimates
  for light traveling in air to our approximate estimates for light
  traveling in a vacuum.

As part of his study, Michelson assessed the various potential sources
of error, and provided his best-guess for the error in his
speed-of-light estimate. These values are provided in
`LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently,
Michelson believed the true value of the speed-of-light probably lay
between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and
`LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of
  that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does
not know the error. The best they can do is quantify their degree of
uncertainty. We will learn some means of quantifying uncertainty in this
class, but for many real problems uncertainty includes some amount of
human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!
df_q3 <-
  tibble(
    Error = abs(LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON),
    Uncertainty = LIGHTSPEED_PM
  ) %>%
  mutate(
    Ratio = Error / Uncertainty
  )

df_q3
```

**Observations**:

- Is Michelson's estimate of the error (his uncertainty) greater or less
  than the true error?

  - At first, 51 km/s sounded like a fairly large uncertainty to me
    especially compared with how precise the speed of light measurement
    seems

  - But once I compared it to the actual difference from the modern
    value, the true error is about 151.5 km/s

  - Michelson's uncertainty was actually much smaller than the error
    that occurred

- Make a quantitative comparison between Michelson's uncertainty and his
  error.

  - The actual error is about 2.97 times Michelson's stated uncertainty

  - His uncertainty suggested that the true value would probably be
    between 299893 and 299995 km/s but the modern value is 299792.458
    km/s which is outside that range

  - This makes me think that there were probably sources of uncertainty
    in the experiment that his estimate did not fully account for

The following plot shows all of Michelson's data as a [control
chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot
is common in manufacturing, where it is used to help determine if a
manufacturing process is under [statistical
control](https://en.wikipedia.org/wiki/Statistical_process_control).
Each dot is one of Michelson's measurements, and the grey line connects
the mean taken for each day. The same plot also shows simulated data
using a probability model. We'll get into statistics later in the
course; for now, let's focus on understanding what real and simulated
data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:

- Similarities

  - The two plots looked fairly similar to me because both have about
    the same center and a similar amount of vertical spread. That makes
    sense after looking back at the code since the simulated data were
    created using the mean and standard deviation of the real
    measurements.

  -  Both also have individual measurements above and below Michelson's
    estimate so neither dataset just forms one tight line around the
    mean.

-  Differences

  - Once I started paying more attention to how the points are grouped
    by date the real data looked less random to me. There are several
    days where a bunch of measurements are high together and other days
    where a bunch are lower together.

  -  The simulated measurements seem more randomly scattered from day to
    day. Their daily average moves around but I do not see the same kind
    of repeated clustering that I see in the real measurements.

  - inititally I thought the difference between the real and simulated
    plots was mainly about having a few unusual points. But I think the
    more interesting difference is actually the day to day structure.
    This makes me wonder if some part of the experimental conditions was
    changing between days instead of every measurement just being a
    random variation around one fixed value.

### **q5** You have access to a few other variables. Construct a **at least three** visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

Visualization of temperature

```{r}
df_q2 %>%
  ggplot(aes(Temp, VelocityVacuum)) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed"
  ) +
  geom_point() +
  theme_minimal() +
  labs(
    x = "Temperature (F)",
    y = "Velocity in Vacuum (km/s)",
    title = "Velocity vs. Temperature"
  )
```

**Observations**:

- first, I did not see a very obvious relationship because the points
  are pretty spread out at almost every temperature. But looking more
  carefully the lower temperature measurements seem to include more of
  the lower velocity values while the measurements at higher
  temperatures are generally a little higher.

<!-- -->

-  I would not say temperature explains the whole difference because
  there is still a lot of overlap. But it does look like temperature
  could be related to the measured velocity.

<!-- -->

-  I also think this is interesting because the 92 km/s correction was
  applied equally to every measurement even though the measurements were
  taken at different temperatures. That makes me wonder whether using
  one correction for all of the measurements could leave some
  temperature-related error behind.

Visualization of distinctness

```{r}
df_q2 %>%
  ggplot(aes(Distinctness, VelocityVacuum)) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed"
  ) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    x = "Distinctness",
    y = "Velocity in Vacuum (km/s)",
    title = "Velocity vs. Image Distinctness"
  )
```

**Observations**

- I expected there to be a steady pattern where better distinctness gave
  higher measurements. But the plot does not really show that.
  Distinctness 2 and 3 look pretty similar, while distinctness 1 is the
  group that seems lower.

<!-- -->

-  This matches what I saw in Q1. It makes me think the important
  difference might not be between every level of distinctness but more
  specifically between the poor measurements and the fair/good
  measurements.

<!-- -->

-  There is still quite a bit of overlap between the groups though so I
  do not think distinctness alone explains Michelson's error.

Visualization of date

```{r}
df_q2 %>%
  ggplot(aes(Date, VelocityVacuum)) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed"
  ) +
  geom_line(
    data = . %>%
      group_by(Date) %>%
      summarize(
        VelocityVacuum = mean(VelocityVacuum)
      ),
    color = "grey50"
  ) +
  geom_point(
    size = 0.8
  ) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity in Vacuum (km/s)",
    title = "Velocity vs. Date"
  )
```

- 

  **Observations**

- I was mostly looking for a simple trend where the measurements either
  increased or decreased over time. There might be a slight decrease
  overall, but I think the more noticeable thing is how much the daily
  averages move around.

-  Some days have several measurements that are all relatively high,
  while other days have measurements that are mostly lower. That seems
  similar to what I noticed in Q4.

-  This makes me think something about the experimental conditions may
  have been changing from day to day. I cannot tell from this plot
  exactly what that factor was, but the measurements do not look like
  completely independent random measurements around one constant value.

## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of
  Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
  (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical
  Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
  (1986)
- [3] BYU video about a [Fizeau-Foucault
  apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to
  what Michelson used.
